---
title: "p8105_hw2_jmm2511"
author: "Jessie Merrifield"
date: "2023-10-03"
output: github_document
---

```{r setup, message=FALSE}
library(tidyverse)
library(readxl)
```

## Question 1

Using datasets: pols-month.csv, unemployment.csv, and snp.csv.

*Goal*: 
merge these into a single data frame using year and month as keys across datasets.


1. Clean dataset pols_month

 Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r clean pols_month}

#pols_month <- read_csv(file #="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/pols-month.csv") %>% 
 # separate(mon, into=c("year", "month", "day"), convert = TRUE) %>% 
  #mutate(month = month.name[as.numeric(month)])%>% 
  #mutate(president=ifelse(prez_dem>0,"dem", "gop")) %>% 
  #select(-day, -prez_dem, -prez_gop)

#print(class(pols_month$month))

month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols_month = 
  read_csv("C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez")) 
 
```

2. Clean dataset snp

clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns

```{r clean snp}

#snp <- read_csv(file ="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/snp.csv")|>
 # separate(date, into=c("month", "day", "year"), convert = TRUE) |>
  #relocate(year, .before=month)|>
  #mutate(
   # year = if_else(year > 2023, year - 100, year))|> 
  #select(year, month, close) 

snp = 
  read_csv(
    "C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/snp.csv",
    col_types = cols(date = col_date(format = "%m/%d/%y"))) |>
  separate(date, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    year = if_else(year > 2023, year - 100, year)) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, close) 

```


3. Clean dataset unemployment

tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values

```{r clean unemployment}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

unemployment = 
  read_csv("C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)

```


Finally, merge the datasets




- Explain briefly what each dataset contained

- describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

Notice that there are some `NA` values in the `close` and `unemployment` variables, which indicate that the value of these variables is missing at those locations.




## Question 2

Read and clean Mr. Trash Wheel dataset

```{r clean Mr. Trash Wheel}
mr_trashwheel<-
  read_excel("C:/Users/jmoon/Downloads/202309 Trash Wheel Collection Data (1).xlsx",
             sheet="Mr. Trash Wheel",
             range = "A2:N586") %>% 
  janitor::clean_names() %>% 
  mutate(homes_powered=(weight_tons*(500/30))) %>%
  mutate(source="mr") %>% 
  mutate(year=as.numeric(year))

#584 observations of 15 variables

```


Use a similar process to import, clean, and organize the data for Professor Trash Wheel

```{r clean Professor Trash Wheel}
prof_trashwheel<-
  read_excel("C:/Users/jmoon/Downloads/202309 Trash Wheel Collection Data (1).xlsx",
             sheet="Professor Trash Wheel",
             range = "A2:M108") %>% 
  janitor::clean_names() %>% 
  mutate(homes_powered=(weight_tons*(500/30)))%>% 
  mutate(source="prof")

#106 observations of 14 variables

#216.26 tons total of trash collected

```


Use a similar process to import, clean, and organize the data for Gwynnda,

```{r Gwynnda Trash Wheel}
g_trashwheel<-
  read_excel("C:/Users/jmoon/Downloads/202309 Trash Wheel Collection Data (1).xlsx",
             sheet="Gwynnda Trash Wheel",
             range = "A2:L157") %>% 
  janitor::clean_names() %>% 
  mutate(homes_powered=(weight_tons*(500/30)))%>% 
  mutate(source="gwynnda")

#155 observations of 13 variables

filter(g_trashwheel,year==2021, month=="July")

#16,3000 cigarette butts from July 2021

```


combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset. 

```{r}
trashwheel_tidy = 
  bind_rows(mr_trashwheel, prof_trashwheel, g_trashwheel) %>% 
  relocate(source) %>% 
  separate(date, into=c("year1", "month1", "day"), convert = TRUE) %>% 
  select(-year1,-month1) %>% 
  relocate(day, .before=year)

#used "source" variable to keep track of which data was from which dataset

#I chose to separate out the day because the date value felt redundant

#there are 845 observations with 15 variables

```


*Write a paragraph about these data; you are encouraged to use inline R.*

There were many more observations from mr_trashwheel than from prof_trashwheel or 
gwynnda_trashwheel, while prof_trashwheel and gwynnda_trashwheel were similar. In this data, Mr. Trash Wheel ran from 2014-2023, Professor Trash Wheel ran from 2017-2023, and Gwynnda Trash Wheel ran from 2021-2023. Gwynnda Trashwheel found no glass bottles or sports balls and 
Professor Trash Wheel found no sports balls. Homes powered ranged from around 10 to around 94.


## Question 3

Import, clean, and tidy the dataset of baseline demographics. 


- Discuss important steps in the import process and relevant features of the dataset. 

```{r read and clean baseline}
baseline<-
  read.csv("C:/Users/jmoon/OneDrive/Desktop/data_mci/MCI_baseline.csv",
           skip = 1) %>% 
#used skip to ensure the first row is headings 
  janitor::clean_names() %>% 
#cleaning names
  mutate(sex=recode(sex, "1"="male", "0"="female")) %>% 
  mutate(apoe4=recode(apoe4, "1"="carrier", "0"="non-carrier")) %>% 
  filter(age_at_onset !=".")
#removed missing age_at_onset values 


print(class(baseline$sex))
print(class(baseline$apoe4))

#sex and APOE4 carrier status were both encoded as integers-->changed to characters


```

*How many participants were recruited, and of these how many develop MCI?*

483 participants were recruited and 97 developed MCI

```{r results='hide'}
print(class(baseline$age_at_onset))

vector_age<-as.numeric(baseline$age_at_onset)
print(mean(vector_age))

filter(baseline, sex=="female")
#46 observations

filter(baseline, sex=="female",apoe4=="carrier")
#30 observations

30/46

```

*What is the average baseline age?* 

The average baseline age is 70.26289.

*What proportion of women in the study are APOE4 carriers?)*

0.6521739

65.22% of women in the study are APOE4 carriers--of those that developed MCI




Import, clean, and tidy the dataset of longitudinally observed biomarker values; 

*comment on the steps on the import process and the features of the dataset.*


```{r}
biomarkers<-read.csv("C:/Users/jmoon/OneDrive/Desktop/data_mci/mci_amyloid.csv",
                     skip=1) %>% 
  janitor::clean_names() %>% 
  rename(id=study_id)
#This dataset is formatted wide, but I do not wish to change it to longer as that would affect the join options with the baseline dataset.
```


*Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings.*

Using join and anti_join to do this

```{r}
base_bio = 
  left_join(baseline, biomarkers, by = "id")
#left_joining to baseline because there are fewer participants in the baseline dataset.
#those who are in the biomarkers dataset but are not in the baseline dataset will not be joined


```



```{r}
anti_joined_data<-anti_join(baseline, biomarkers, by="id")
```


After filtering out those participants that never develop MCI, there are many participants
that only appear in the biomarkers dataset. However, there are three other participants that only appear in the baseline dataset. 


```{r}
baseline_filtered<-filter(baseline, id !="14", id !="49", id !="268")

#filtering out the participant id's of those in the baseline dataset but not the biomarkers dataset

```


Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained


```{r}
base_bio_final = 
  left_join(baseline_filtered, biomarkers, by = "id")

```


*briefly describe the resulting dataset;*

There are 94 observations of 11 variables in this dataset. The "current ages" of participants 
range from 58.1 years to 71.6 years. The youngest age at onset was 61.2 and the oldest age 
at onset was 77.2 years. There are 60 carriers and 34 non-carriers and there are 44 females 
and 50 males. 

```{r}
count(base_bio_final, sex)
count(base_bio_final, apoe4)
```


export the result as a CSV to your data directory.

```{r}
write.csv(base_bio_final, "C:/Users/jmoon/OneDrive/Desktop/data_mci/base_bio.csv", row.names=FALSE)
```




