---
title: "p8105_hw2_jmm2511"
author: "Jessie Merrifield"
date: "2023-10-03"
output: github_document
---

```{r setup, message=FALSE}
library(tidyverse)

```

## Question 1

Using datasets: pols-month.csv, unemployment.csv, and snp.csv.

*Goal*: 
merge these into a single data frame using year and month as keys across datasets.

```{r read in, message=FALSE}
pols_month <- read_csv(file ="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/pols-month.csv")

unemployment <- read_csv(file ="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/unemployment.csv")

snp <- read_csv(file ="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/snp.csv")

```

1. Clean dataset pols_month

 Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r clean pols_month}

pols_month <- read_csv(file ="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into=c("year", "month", "day")) |>
  mutate("month"=month.name[month])

?separate
?month.name

```

2. Clean dataset snp

clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns

```{r clean snp}
snp <- read_csv
(file ="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/snp.csv")|>
  separate(mon, into=c("year", "month", "day")) |>
  mutate("month"=month.name[month])
```


3. Clean dataset unemployment

tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values

```{r clean unemployment}

```


Finally, merge the datasets

Join the datasets by merging snp into pols, and merging unemployment into the result

```{r join Q1}

```

*Short paragraph about the datasets in Q1:*

- Explain briefly what each dataset contained
- describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

## Question 2

Read and clean Mr. Trash Wheel dataset

- specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
- use reasonable variable names
- omit rows that do not include dumpster-specific data

Update the data to include a new homes_powered variable based on this calculation

```{r}

```


Use a similar process to import, clean, and organize the data for Professor Trash Wheel

```{r}

```


Use a similar process to import, clean, and organize the data for Gwynnda,

```{r}

```


combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to all datasets before combining.

```{r}

```


*Write a paragraph about these data; you are encouraged to use inline R.*

- note the number of observations in the resulting dataset
- give examples of key variables. 
- what was the total weight of trash collected by Professor Trash Wheel?
- What was the total number of cigarette butts collected by Gwynnda in July of 2021?


## Question 3

Import, clean, and tidy the dataset of baseline demographics. 

- Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric)
- remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline).
- Discuss important steps in the import process and relevant features of the dataset. 

```{r}

```

*How many participants were recruited, and of these how many develop MCI?*

*What is the average baseline age?* 

*What proportion of women in the study are APOE4 carriers?)*



import, clean, and tidy the dataset of longitudinally observed biomarker values; 

*comment on the steps on the import process and the features of the dataset.*

```{r}

```


*Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings.*



Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained

```{r}

```


*briefly describe the resulting dataset;*

export the result as a CSV to your data directory.

```{r}

```




