---
title: "p8105_hw2_jmm2511"
author: "Jessie Merrifield"
date: "2023-10-03"
output: github_document
---

```{r setup, message=FALSE}
library(tidyverse)

```

## Question 1

Using datasets: pols-month.csv, unemployment.csv, and snp.csv.

*Goal* merge these into a single data frame using year and month as keys across datasets.

```{r read in, message=FALSE}
pols_month <- read_csv(file ="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/pols-month.csv")

unemployment <- read_csv(file ="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/unemployment.csv")

snp <- read_csv(file ="C:/Users/jmoon/OneDrive/Desktop/fivethirtyeight_datasets/snp.csv")

```

1. Clean dataset pols_month

 Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r clean pols_month}

```

2. Clean dataset snp

clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns

```{r clean snp}

```


3. Clean dataset unemployment

tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values

```{r clean unemployment}

```


Finally, merge the datasets

Join the datasets by merging snp into pols, and merging unemployment into the result

```{r join Q1}

```

Short paragraph about the datasets in Q1:

Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

## Question 2


```{r}

```


## Question 3


```{r}

```